NumLin: Linear Types for Numerical Libraries

### 2018-10-21 write-up/paper/lang_and_examples.tex
Add matrix squaring example

The more examples, the better. In fact, it's actually impossible to
have too many examples -- it's okay (indeed, desirable) to spend 5-6
pages on examples.
\begin{itemize}

    \item Simple: factorial, shows recursion, ! annotations etc.

    \item Less simple: summing over an array, indexing and safety

    \item Medium: one-dimensioal convolution, permissions

    \item Harder: linear regression, pattern-matching and apparent non-linearity

    \item Harder: L1 norm-minimisation, some frees, re-using memory

    \item Big finish: Kalman filter

@@

The condition for an \highl{if} may or may not use linear values (here, with
\highl{x < 0 || x = 0}, it does not). Any linear values used by the conditional
would not be in scope in either branch of the \highl{if}-expression.  Both
branches use \texttt{x} differently: one ignores it completely and the other
uses it twice.

@@

Second is \highl{('x) (row : 'x arr)} which is an array with a
universally-quantified fractional permission. In particular, this means the
body of the function cannot mutate or free the input array, only read from it.
If the programmer did try to mutate or free \texttt{row}, then they would get a
helpful error message (Figure~\ref{fig:lang_errormsg}).

@@

Alongside taking a \highl{row : 'x arr}, the function also returns an array
with exactly the same fractional permission as the \texttt{row} (which can only
be \texttt{row}).  This is necessary because of linearity: for the caller, the
original array passed in as an argument would be out of scope for the rest of
the expression, so it needs to be returned and then rebound to be used for the
rest of the function.

An example of this consuming and re-binding is in \highl{let (row, !x1) =
row[i]}. Indexing is implemented as a primitive \highl{get: 'x. 'x arr --o
!int --o 'x arr * !elt}. Although fractional permissions can be passed around
explicitly  (as done in the recursive call), they can also be
\emph{automatically inferred at call sites}, \highl{row[i] == get _ row i}
takes advantage of this convenience.

@@

What's implemented in Figure~\ref{fig:lang_oned_conv} is a \emph{simplified}
version of this idea, so as to not distract from the features of \lang. The
simplifications are:
\begin{itemize}
    \item the kernel has a fixed length (of 3), as such, only the value of
        \highl{write[i-1]} prior to modification in the previous iteration
        needs to be carried forward using \highl{x0}
    \item \highl{write} is assumed to have length \highl{n+1}
    \item \texttt{i}'s initial value is assumed to be \texttt{1}
    \item \highl{x0}'s initial value is assumed to be \highl{write[0]}
    \item the first and last values of \texttt{write} are ignored.
\end{itemize}

@@

@@

\emph{The most pertinent aspect of \lang's type system are the types of its
primitives.} While the types of operations such as \texttt{get} and
\texttt{set} might be borderline obvious, they types of BLAS/LAPACK routines
become an \emph{incredibly useful, automated check for using the API correctly.}

Figure~\ref{fig:lang_lin_reg} shows a \emph{non-recursive} function declaration (the
return type is inferred). Since we would like to be able to use a function like
\texttt{lin\_reg} more than once, it is marked with a \texttt{!} annotation
(which also ensure it captures no linear values from the surrounding
environment).

We wish to compute $\hat\beta = \mathbf{(X^T X)^{-1} X^T y}$. First, we extract
the dimensions of matrix \texttt{x}. Then, we say we would like a new matrix, of
dimension $m \times 1$, which contains the result of $\mathbf{X^T y}$. The line
\highl{let xy <- new (m,1) [| x^T * y |]} is syntactic sugar for first creating
a new matrix \highl{let xy = matrix m 1 in} and then storing the result of
\highl{(x, true (* transp'd *))} multiplied with \highl{(y, false (* not
transp'd *))} into \texttt{xy}: \highl{let (xy, (x, y)) = gemm 1.  _ (x,
true) _ (y, false) 0. xy}, appropriately re-binding the identifiers.  The line
\highl{let x_T_x <- new (m,m) [| x^T * x |]}, works similarly, but uses a more
appropriate, \highl{syrk true 1. x 0. x_T_x} BLAS call.

By using some simple pattern-matching and syntactic sugar, we can:
\begin{itemize}
    \item write normal-looking, apparently non-linear code
    \item use matrix expressions directly and have an appropriate, efficient
        call to a BLAS/LAPACK routine inserted
    \item retain the safety of linear types with fractional permissions by
        having the compiler statically enforce the aliasing and read/write rules
        implicitly assumed by BLAS/LAPACK routines.
\end{itemize}

@@

Figure~\ref{fig:lang_l1_norm_min} shows even more pattern-matching. Patterns of
the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are desugared to
\texttt{gemm} calls. Primitives like \highl{transpose: 'x. 'x mat --o 'x mat *
z mat} and \highl{eye: !int --o z mat} allocate new matrices;
\texttt{transpose} returns the transpose of a given matrix and \texttt{eye}
returns a $k \times k$ identity matrix.

We also see our first example of re-using memory for different matrices: like
with \texttt{to\_del} and \texttt{posv} in the previous example, we do not need
the value stored in \texttt{tmp\_5\_5} after the call to \texttt{gesv}. However,
we can use the memory much later to store \texttt{answer} with \highl{let
answer <- [| 0. * tmp_5_5 + q_inv_u * inv_u_T |]}. Again, thanks to linearity,
the identifiers \texttt{q}, \texttt{tmp\_5\_5} are out of scope by the time
\texttt{answer} is bound. Although during execution, all three refer to the
same piece of memory, logically they represent different values throughout the
computation.

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/kalman.lt}
    \caption{Kalman filter: see Figure~\ref{fig:maths_kalman} for the
        equations this code implements.}\label{fig:lang_kalman}
\end{figure}

\begin{figure}[t]
    \begin{align*}
        \mu' &= \mu + \Sigma H^T (R + H \Sigma H^T)^{-1} (H \mu - \textrm{data})\\
        \Sigma' &= \Sigma ( I - H^T (R + H \Sigma H^T)^{-1} H \Sigma )
    \end{align*}
    \caption{Kalman filter equations (credit:
    \href{http://matthewrocklin.com/blog/work/2012/11/24/Kalman-Filter}{matthewrocklin.com}).}\label{fig:maths_kalman}
\end{figure}

Figure~\ref{fig:lang_kalman} shows a \lang\ implementation of a Kalman filter
(equations in Figure~\ref{fig:maths_kalman}). A few new features and techniques
are used in this implementation:
\begin{itemize}

    \item \texttt{sym} annotations in matrix expressions: when this is used, a 
        call to \texttt{symm} (the equivalent of \texttt{gemm} but for
        symmetric matrices so that only half the operations are performed) is
        inserted (instead of one to \texttt{gemm})

@@

    \item a lot of memory re-use; the following sets of identifiers alias each other:
        \begin{itemize}
            \item \texttt{r\_1} and \texttt{r\_2}
            \item \texttt{data\_1}, \texttt{data\_2} and \texttt{sol\_data}
            \item \texttt{new\_h} and \texttt{sol\_h}
            \item \texttt{h\_sol\_h}, \texttt{sigma\_copy} and \texttt{new\_sigma}
            \item \texttt{mu\_copy} and \texttt{new\_mu}
        \end{itemize}

@@

>>>>

@@

The condition for an \highl{if} may or may not use linear values (here, with
\highl{x < 0 || x = 0}, it does not). Any linear values used by the condition
would not be in scope in either branch of the \highl{if}-expression.  Both
branches use \texttt{x} differently: one ignores it completely and the other
uses it twice.

@@

Second is \highl{('x) (row: 'x arr)} which is an array with a
universally-quantified fractional permission. In particular, this means the
body of the function cannot mutate or free the input array, only read from it.
If the programmer did try to mutate or free \texttt{row}, then they would get a
helpful error message (Figure~\ref{fig:lang_errormsg}).

@@

Alongside taking a \highl{row: 'x arr}, the function also returns an array
with exactly the same fractional permission as the \texttt{row} (which can only
be \texttt{row}).  This is necessary because of linearity: for the caller, the
original array passed in as an argument would be out of scope for the rest of
the expression, so it needs to be returned and then rebound to be used for the
rest of the function.

An example of this consuming and re-binding is in \highl{let (row, !x1) =
row[i]}. Indexing is implemented as a primitive \highl{get: 'x. 'x arr --o
!int --o 'x arr * !elt}. Although fractional permissions can be passed around
explicitly  (as done in the recursive call), they can also be
\emph{automatically inferred at call sites}: \highl{row[i] == get _ row i}
takes advantage of this convenience.

@@

What's implemented in Figure~\ref{fig:lang_oned_conv} is a \emph{simplified}
version of this idea, so as to not distract from the features of \lang. The
simplifications are:
\begin{itemize}
    \item the kernel has a length 3, so only the value of \highl{write[i-1]}
        (prior to modification in the previous iteration) needs to be carried
        forward using \highl{x0}
    \item \highl{write} is assumed to have length \highl{n+1}
    \item \texttt{i}'s initial value is assumed to be \texttt{1}
    \item \highl{x0}'s initial value is assumed to be \highl{write[0]}
    \item the first and last values of \texttt{write} are ignored.
\end{itemize}

@@

\subsubsection{Squaring a Matrix}

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/square.lt}
    \caption{Linear regression (OLS): $\hat\beta =
        \mathbf{(X^T X)^{-1} X^T y}$}\label{fig:lang_square}
\end{figure}

\emph{The most pertinent aspect of \lang\ is the types of its primitives.}
While the types of operations such as \texttt{get} and \texttt{set} might be
borderline obvious, the types of BLAS/LAPACK routines become an
\emph{incredibly useful, automated check for using the API correctly.}

Figure~\ref{fig:lang_square} shows how a linearly-typed matrix squaring
function may be written in \lang. It is a \emph{non-recursive} function
declaration (the return type is inferred). Since we would like to be able to
use a function like \texttt{square} more than once, it is marked with a
\texttt{!} annotation (which also ensures it captures no linear values from the
surrounding environment).

To square a matrix, first, we extract the dimensions of the argument
\texttt{x}. Then, because we need to use \texttt{x} twice (so that we can
multiply it by itself) but linearity only allows one use, we use \highl{shareM:
'x. 'x mat --o 'x s mat * 'x s mat} to split the permission \texttt{'x} (which
represents $2^{-x}$) into two halves (\texttt{'x s}, which represents $2^{-(x+1)}$).

Even if \texttt{x} had type \texttt{z mat}, sharing it now enforces the
assumption of all BLAS/LAPACK routines that any matrix which is written to
(which, in \lang, is always of type \texttt{z mat}) does not alias any other
matrix in scope. So if we did try to use one of the aliases in mutating way,
the expression would not type check, and we would get an error similar to the
one in Figure~\ref{fig:lang_errormsg}.

The line \highl{let answer <- new (m,n) [| x1 * x2 |]} is syntactic sugar for
first creating a new $m \times n$ matrix (\highl{let answer = matrix m n}) and
then storing the result of the multiplication in it (\highl{let ((x1, x2),
answer) = gemm 1. _ (x1, false) _ (x2, false) 0. answer}). \highl{false} means
the matrix should not be accessed with indices transposed.

By using some simple pattern-matching and syntactic sugar, we can:
\begin{itemize}
    \item write normal-looking, apparently non-linear code
    \item use matrix expressions directly and have a call to an efficient
        call to a BLAS/LAPACK routine inserted with appropriate re-bindings
    \item retain the safety of linear types with fractional permissions by
        having the compiler statically enforce the aliasing and read/write rules
        implicitly assumed by BLAS/LAPACK routines.
\end{itemize}

@@

In Figure~\ref{fig:lang_lin_reg}, we wish to compute $\hat\beta = \mathbf{(X^T
X)^{-1} X^T y}$. To do that, first, we extract the dimensions of matrix
\texttt{x}. Then, we say we would like \texttt{xy} to be a new matrix, of
dimension $m \times 1$, which contains the result of $\mathbf{X^T y}$ (using
syntactic sugar for \texttt{matrix} and \texttt{gemm} calls similar to that
used in Figure~\ref{fig:lang_square}, with a \highl{^T} annotation on
\texttt{x} to set \texttt{x}'s `transpose indices'-flag to \highl{true}).

However, the line \highl{let x_T_x <- new (m,m) [| x^T * x |]}, works for a
slightly different reason: that pattern is matched to a BLAS call to
(\highl{syrk true 1.  x 0. x_T_x}), which only uses \texttt{x} once. Hence
\texttt{x} can appear \emph{twice} in the \emph{pattern} without any calls to
\texttt{share}.

@@

Figure~\ref{fig:lang_l1_norm_min} shows even more pattern-matching. Patterns of
the form \highl{let <id> <- [| beta * c + alpha * a * b |]} are also desugared to
\texttt{gemm} calls. Primitives like \highl{transpose: 'x. 'x mat --o 'x mat *
z mat} and \highl{eye: !int --o z mat} allocate new matrices;
\texttt{transpose} returns the transpose of a given matrix and \texttt{eye k}
evaluates to a $k \times k$ identity matrix.

We also see our first example of re-using memory for different matrices: like
with \texttt{to\_del} and \texttt{posv} in the previous example, we do not need
the value stored in \texttt{tmp\_5\_5} after the call to \texttt{gesv} (a
primitive similar to \texttt{posv} but for a non-symmetric first argument).
However, we can re-use its memory much later to store \texttt{answer} with
\highl{let answer <- [| 0. * tmp_5_5 + q_inv_u * inv_u_T |]}. Again, thanks to
linearity, the identifiers \texttt{q} and \texttt{tmp\_5\_5} are out of scope
by the time \texttt{answer} is bound. Although during execution, all three
refer to the same piece of memory, logically they represent different values
throughout the computation.

@@

\begin{figure}[t]
    \centering
    \inputminted[fontsize=\small]{ocaml}{../../test/examples/kalman.lt}
    \caption{Kalman filter: see Figure~\ref{fig:kalman_eqns} for the
        equations this code implements. Line numbers in comments refer to
        equivalent lines in a C implementation.}\label{fig:lang_kalman}
\end{figure}

\begin{figure}[t]
    \begin{align*}
        \mu' &= \mu + \Sigma H^T (R + H \Sigma H^T)^{-1} (H \mu - \textrm{data})\\
        \Sigma' &= \Sigma ( I - H^T (R + H \Sigma H^T)^{-1} H \Sigma )
    \end{align*}
    \caption{Kalman filter equations (credit:
    \href{http://matthewrocklin.com/blog/work/2012/11/24/Kalman-Filter}{matthewrocklin.com}).}\label{fig:kalman_eqns}
\end{figure}

Figure~\ref{fig:lang_kalman} shows a \lang\ implementation of a Kalman filter
(equations in Figure~\ref{fig:kalman_eqns}). A few new features and techniques
are used in this implementation:
\begin{itemize}

    \item \texttt{sym} annotations in matrix expressions: when this is used, a 
        call to \texttt{symm} (the equivalent of \texttt{gemm} but for
        symmetric matrices so that only half the operations are performed) is
        inserted

@@

    \item a lot of memory re-use; the following sets of identifiers alias each other:
        \begin{itemize}
            \item \texttt{r\_1} and \texttt{r\_2}
            \item \texttt{data\_1}, \texttt{data\_2} and \texttt{sol\_data}
            \item \texttt{new\_h} and \texttt{sol\_h}
            \item \texttt{h\_sol\_h}, \texttt{sigma\_copy} and \texttt{new\_sigma}
            \item \texttt{mu\_copy} and \texttt{new\_mu}.
        \end{itemize}

@@

With expressions as complex as those in Figure~\ref{fig:kalman_eqns}, it can
get very difficult to ensure memory is not being re-used incorrectly or re-used
enough whilst also enusring the aliasing assumptions for the BLAS/LAPACK
routine calls are guaranteed. However, using \lang\ makes incorrect aliasing
and memory re-use impossible. Additionally, \lang\ makes it very easy to spot
where memory is not being re-used enough \emph{precisely because} of its
explicit handling of matrix allocations, copies and frees.
